Conformal Counterfactual Explanations

### 2023-02-12 dev/proposal.qmd
proposal

@@

Most state-of-the-art approaches to generating Counterfactual Explanations (CE) rely on gradient descent in the feature space. The key idea is to perturb inputs $x\in\mathcal{X}$ into a black-box model $f: \mathcal{X} \mapsto \mathcal{Y}$ in order to change the model output $f(x)$ to some pre-specified target value $t\in\mathcal{Y}$. Formally, this boils down to defining some loss function $\ell(f(x),t)$ and taking gradient steps in the minimizing direction. The so generated counterfactuals are considered valid as soon as the predicted label matches the target label. A stripped down counterfactual is therefore little different from Adversarial Examples [@goodfellow2014explaining].

@@

@@

>>>>

```{julia}
using CounterfactualExplanations
using CounterfactualExplanations.Data: load_mnist
using CounterfactualExplanations.Models: load_mnist_mlp
using Images
using MLDatasets
using MLDatasets: convert2image
using Plots
```

@@

Most state-of-the-art approaches to generating Counterfactual Explanations (CE) rely on gradient descent in the feature space. The key idea is to perturb inputs $x\in\mathcal{X}$ into a black-box model $f: \mathcal{X} \mapsto \mathcal{Y}$ in order to change the model output $f(x)$ to some pre-specified target value $t\in\mathcal{Y}$. Formally, this boils down to defining some loss function $\ell(f(x),t)$ and taking gradient steps in the minimizing direction. The so generated counterfactuals are considered valid as soon as the predicted label matches the target label. A stripped down counterfactual explanation is therefore little different from an adversarial example.

@@

```{julia}
# Data:
counterfactual_data = load_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()

# Target:
factual_label = 8
x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x)[1]

# Search:
n_ce = 3
generator = GenericGenerator()
ces = generate_counterfactual(x, target, counterfactual_data, M, generator; num_counterfactuals=n_ce)
```

```{julia}
image_size = 300
p1 = plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(image_size, image_size),
    title="Factual"
)
plts = [p1]

for i in eachindex(ces)
    ce = ces[i]
    plt = plot(
        convert2image(MNIST, reshape(CounterfactualExplanations.counterfactual(ce),28,28)),
        axis=nothing, 
        size=(image_size, image_size),
        title="Counterfactual $i"
    )
    plts = [plts..., plt]
end

plt = plot(plts...; size=(image_size * (n_ce + 1),image_size), layout=(1,(n_ce + 1)))
```

The crucial difference between adversarial examples and counterfactuals is one of intent. While adversarial examples are typically intened to go unnoticed, counterfactuals in the context of Explainable AI are generally sought to be "plausible" or "realistic". To fulfill this latter goal, researchers have come up with a myriad of ways. @joshi2019realistic were among the first to suggest that instead of searching counterfactuals in the feature space, we can instead traverse a latent embedding learned by a surrogate generative model. This ensures that the generated counterfactuals comply with the (learned) data-generating process (DGB). Similarly, @poyiadzi2020face use density ...

@@

::: {#prp-surrogate}

## Avoid Surrogates

Since we are in the business of explaining a black-box model, the task of learning realistic representations of the data should not be reallocated from the model itself to some surrogate model.

:::
